Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-10 VegOzone.tex
Spell-checked

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This convention aims to develope an effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this the concept of critical levels (CLs) has been developed and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep{ICP:MappingManual2017}. The CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018 (Section~\ref{sec:do3se}). In this end, we erive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally bio-geochemical cycling.

@@

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on conditions during the 2018/19 growing season at the NIBIO operated Environment Centre Svanhovd. We also set these into a climatological and regional context. To this end, we collected atmospheric and ozone monitoring data taken at Svanhovd and additional ozone monitoring data from observation sites in northern Finland and Sweden for cross-evaluation. Varring update frequencies and length of available data sets made it necessary to fetch the data from various sources (further details below). Svanhovd synonymously refers to the atmospheric monitoring site Svanvik/Pasvik throughout the paper.\\

@@

All relevant data for the growing season 2018/19 are shown as timeseries in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations measured at $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions. To this end, we compute multi-annual mean (referred to as climatology in the following) of environmental key variables (temperature, precipitation, global irradiance) and gound-level \chem{[O_3]} at Svanhovd. For the growing seasons 2018/19, we evaluate the statistical significance of divergances from the norm in these variables (referred to as anomalies). We perform a bias correction accounting for the increase in tropospheric background \chem{[O_3]} and cross-calibrate the derived climatology for Svanvik with long-term observations of \chem{[O_3]} at Esrange and Pallas.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we derive an ozone climatology (\chem{\left<[O_3]\right>}) for Svanvik and for northern Fennoscadia. To this end, we first calculate a pairwise correlation between the different stations. The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Appendix Fig.~\ref{fig:density_distribution}a, b). Therefore, we select these data to compute the generallized ozone climatology for northern Fennoscandia. The correlation between Svanvik with any of the active monitoring sites in 2018 (Esrange and Pallas) is rather low, 0.4 and 0.6, respectively. Thus, conditions at Svanvik cannot directly be represented by measurements taken at either Esrange or Pallas.

@@

The generalized northern Fennoscandic ozone climatology peaks at $48\,\unit{ppb}$  in late April (day of the year (doy) $100-120$) and declines throughout the growing season until a minimum is reached in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. The most pronounced increase in \chem{\left<[O_3]\right>} coincidents with rising temperatures and global irradiance. The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} for Svanvik display a similar pattern as the generallized northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier ($100\,\unit{doy}$) and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than the Fennoscandic climatology. This is in line with a general decrease in tropospheric ozone with decreasing altitude. The earlier decline in \chem{\left<[O_3]\right>} may point to an influence of coniferous vegetation on the removal of ozone since \chem{CO_2} uptake by coniferous trees has been observed as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013} compared to an estimated average begin of growing season ($\left<G_\mathrm{start}\right>$) on $(150\pm 14)\,\unit{doy}$ at Svanvik (see Section~\ref{subsec:do3se_parameters} for details). In July--September, ozone is occasionally almost completely depleted during nighttime. This hint to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air. The derived \chem{\left<[O_3]\right>} is likely biased towards lower \chem{[O_3]} due to an increase in northern hemispheric tropospheric background \chem{[O_3]} since the 1990s. To account for this, we derive a bias correction factor by cross-calibrating with the timeseries taken at Esrange and Pallas (Section~\ref{subsec:anomalies}).

@@

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these data, we performed a Reynold's decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone obserations at Esrange and Pallas (compagnion paper). In the following, we will describe the bias-correction.

@@

Similarly, we find a skew towards higher \chem{[O_3]} for Svanvik in 2018, while 2019 follows a normal distribution rather well (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}). In conjunction with the meteorological conditions, we therefore can assume that 2019 had been a normal year and that the annual average anomally in 2019 can be used for the bias correction of the Svanvik climatology (\chem{\left<\Delta[O_3]\right>}~=~$(1.2\pm 5.0)\,\unit{ppb}$). To further strengthen this, we look at trends in ozone observations at Esrange and Pallas and compute the average difference between recent years and the mid 1990s (\chem{\Delta_{present-past}\left<[O_3]\right>}). Present is defined as the average over 2010--2018 and past as the time between the respective beginning of data taking until 1997. We do not find a clear trend. For Esrange \chem{\Delta_{present-past}\left<[O_3]\right>} amounts to $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas to $(-2.5\pm 3.7)\,\unit{ppb}$ \citep[][for comparison]{NILU2003, BER:Ruoho-Airola2015}. The observed trend at Esrange is compatible with our assumed bias correction for the Svanvik ozone climatology.

@@

\section{DO3SE modeling}
\label{sec:do3se}
To investigate the difference between 2018 and 2019 concerning stomatal ozone uptake and to assess which of the environmental factors might be most influential in determining \chem{O_3} damage risk, we model ozone uptake of different natural and semi-natural vegetation using the $\mathrm{DO_3SE}$ model and develop bespoke subarctic parameterizations for common plant species. We the $\mathrm{DO_3SE}$ model and gapp-filling methods are described in Appendix~\ref{appendix:do3se_model}. First, we give an account of the methodology for deriving bespoke parameters. Then, we will present and discuss the modeling results.

@@

The main species or PFTs, we deduced on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station}, are perennial grassland, birch, and pine. Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} abbreviated as MM in the following. We base our parameterization of coniferous trees on boreal Norway spruce from the MM. Birch will be referred to as deciduous tree in the following. A comprehensive list of model parameters can be found in Supplement~Table~S1.
Initial test simulations with the MM PFTs showed unrealistically low $g_\mathrm{sto}$ in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Only during an extended warm period in late July did we find substantial stomatal conductance. In ecological terms this would mean that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance. Therefore, we propose bespoke parameterizations which may describe an actual acclimation of these species to the climatic conditions in the target region.

@@

For the missing ozone data, we applyed a Reynolds's decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with climatology \chem{\left<[O_3]\right>} and anomalies \chem{\Delta[O_3]}. Observed \chem{[O_3]} in 2018 at Pallas served as the reference. The exact procedure will be described in an compagnion paper.

@@

To assess the $G_\mathrm{start}$  and $G_\mathrm{end}$ of the growing season for coniferous trees at Svanhovd in 2018/19, we used the net photosynthesis product ($A_\mathrm{net}$) of MODIS AQUA/TERRA over a $1\times 1\,\unit{km}$ area centered at Svanhovd. MODIS data indicate a higher photosynthetic activity in 2018 than in 2019. As shown in Fig.~\ref{fig:modis_Psn} we fitted a second order polynomal function of the general form
%
\begin{equation}
A_\mathrm{net}(t) =  -m_0\cdot(t-m_1)^2+m_2,
\end{equation}
%
with the form parameters $m_i$ through the data. Numerically, we retrieved the roots as $G_\mathrm{start}$ 122/106 and $G_\mathrm{end}$ 261/274 for 2018 and 2019, respectively.

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV contributed to framing this research article, has performed the on-site observation of vegetation damage induced by ozone, and advision in plant physiological processes. LE has contributed to framing of the this article, $\mathrm{DO_3SE}$ modeling, and development of the PDF-based temperature acclimation. CO has collected species or PFT parameters from literature, performed all $\mathrm{DO_3SE}$ simulations and initial validation of the results. AE contributed with her experience regarding subarctic vegetation in Finnmark. FS contributed with an assessment of the 2018 meteorological conditions. TB gave guidance in a broader research sense. All authors contributed to discussions, writing, and conclusions.} %% this section is mandatory

>>>>

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This convention aims to develop an effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this the concept of critical levels (CLs) has been developed and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep{ICP:MappingManual2017}. The CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive multi-annual means for ground-level ozone concentration and key environmental variables (temperature, precipitation, and global irradiance) based on in situ observations and evaluate the deviation of 2018/19 from the norm. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018 (Section~\ref{sec:do3se}). In this end, we derive bespoke parameterizations for subarctic species to assess systematic uncertainties arising from the use of default generic parameterizations established from less climatically extreme conditions. In Section~\ref{sec:conc}, we will summarize our results and discuss how this research can support improvements in future assessments of pollution risk and more generally bio-biogeochemical cycling.

@@

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on conditions during the 2018/19 growing season at the NIBIO operated Environment Centre Svanhovd. We also set these into a climatological and regional context. To this end, we collected atmospheric and ozone monitoring data taken at Svanhovd and additional ozone monitoring data from observation sites in northern Finland and Sweden for cross-evaluation. Varying update frequencies and length of available data sets made it necessary to fetch the data from various sources (further details below). Svanhovd synonymously refers to the atmospheric monitoring site Svanvik/Pasvik throughout the paper.\\

@@

All relevant data for the growing season 2018/19 are shown as time series in Fig.~\ref{fig:data_svanvik_2018_2019}. Ozone concentrations measured at $2\,\unit{m}$ height above ground are averaged hourly. The hatched areas mark times when no ozone data were recorded. Note, while the downtime during winter was planned, missing data in two weeks of July 2018 (July 9--23) were due to problems in data acquisition.

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions. To this end, we compute multi-annual mean (referred to as climatology in the following) of environmental key variables (temperature, precipitation, global irradiance) and ground-level \chem{[O_3]} at Svanhovd. For the growing seasons 2018/19, we evaluate the statistical significance of divergences from the norm in these variables (referred to as anomalies). We perform a bias correction accounting for the increase in tropospheric background \chem{[O_3]} and cross-calibrate the derived climatology for Svanvik with long-term observations of \chem{[O_3]} at Esrange and Pallas.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we derive an ozone climatology (\chem{\left<[O_3]\right>}) for Svanvik and for northern Fennoscandia. To this end, we first calculate a pairwise correlation between the different stations. The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Appendix Fig.~\ref{fig:density_distribution}a, b). Therefore, we select these data to compute the generalized ozone climatology for northern Fennoscandia. The correlation between Svanvik with any of the active monitoring sites in 2018 (Esrange and Pallas) is rather low, 0.4 and 0.6, respectively. Thus, conditions at Svanvik cannot directly be represented by measurements taken at either Esrange or Pallas.

@@

The generalized northern Fennoscandic ozone climatology peaks at $48\,\unit{ppb}$  in late April (day of the year (doy) $100-120$) and declines throughout the growing season until a minimum is reached in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. The most pronounced increase in \chem{\left<[O_3]\right>} coincident with rising temperatures and global irradiance. The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} for Svanvik display a similar pattern as the generalized northern Fennoscandia climatology (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b). The spring peak occurs slightly earlier ($100\,\unit{doy}$) and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than the Fennoscandic climatology. This is in line with a general decrease in tropospheric ozone with decreasing altitude. The earlier decline in \chem{\left<[O_3]\right>} may point to an influence of coniferous vegetation on the removal of ozone since \chem{CO_2} uptake by coniferous trees has been observed as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013} compared to an estimated average begin of growing season ($\left<G_\mathrm{start}\right>$) on $(150\pm 14)\,\unit{doy}$ at Svanvik (see Section~\ref{subsec:do3se_parameters} for details). In July--September, ozone is occasionally almost completely depleted during nighttime. This hint to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air. The derived \chem{\left<[O_3]\right>} is likely biased towards lower \chem{[O_3]} due to an increase in northern hemispheric tropospheric background \chem{[O_3]} since the 1990s. To account for this, we derive a bias correction factor by cross-calibrating with the time series taken at Esrange and Pallas (Section~\ref{subsec:anomalies}).

@@

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these data, we performed a Reynolds decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone observations at Esrange and Pallas (companion paper). In the following, we will describe the bias-correction.

@@

Similarly, we find a skew towards higher \chem{[O_3]} for Svanvik in 2018, while 2019 follows a normal distribution rather well (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}). In conjunction with the meteorological conditions, we therefore can assume that 2019 had been a normal year and that the annual average anomaly in 2019 can be used for the bias correction of the Svanvik climatology (\chem{\left<\Delta[O_3]\right>}~=~$(1.2\pm 5.0)\,\unit{ppb}$). To further strengthen this, we look at trends in ozone observations at Esrange and Pallas and compute the average difference between recent years and the mid 1990s (\chem{\Delta_{present-past}\left<[O_3]\right>}). Present is defined as the average over 2010--2018 and past as the time between the respective beginning of data taking until 1997. We do not find a clear trend. For Esrange \chem{\Delta_{present-past}\left<[O_3]\right>} amounts to $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas to $(-2.5\pm 3.7)\,\unit{ppb}$ \citep[][for comparison]{NILU2003, BER:Ruoho-Airola2015}. The observed trend at Esrange is compatible with our assumed bias correction for the Svanvik ozone climatology.

@@

\section{DO3SE modeling}
\label{sec:do3se}
To investigate the difference between 2018 and 2019 concerning stomatal ozone uptake and to assess which of the environmental factors might be most influential in determining \chem{O_3} damage risk, we model ozone uptake of different natural and semi-natural vegetation using the $\mathrm{DO_3SE}$ model and develop bespoke subarctic parameterizations for common plant species. We the $\mathrm{DO_3SE}$ model and gap-filling methods are described in Appendix~\ref{appendix:do3se_model}. First, we give an account of the methodology for deriving bespoke parameters. Then, we will present and discuss the modeling results.

@@

The main species or PFTs, we deduced on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station}, are perennial grassland, birch, and pine. Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} abbreviated as MM in the following. We base our parameterization of coniferous trees on Boreal Norway spruce from the MM. Birch will be referred to as deciduous tree in the following. A comprehensive list of model parameters can be found in Supplement~Table~S1.
Initial test simulations with the MM PFTs showed unrealistically low $g_\mathrm{sto}$ in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Only during an extended warm period in late July did we find substantial stomatal conductance. In ecological terms this would mean that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance. Therefore, we propose bespoke parameterizations which may describe an actual acclimation of these species to the climatic conditions in the target region.

@@

For the missing ozone data, we applied a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with climatology \chem{\left<[O_3]\right>} and anomalies \chem{\Delta[O_3]}. Observed \chem{[O_3]} in 2018 at Pallas served as the reference. The exact procedure will be described in an companion paper.

@@

To assess the $G_\mathrm{start}$  and $G_\mathrm{end}$ of the growing season for coniferous trees at Svanhovd in 2018/19, we used the net photosynthesis product ($A_\mathrm{net}$) of MODIS AQUA/TERRA over a $1\times 1\,\unit{km}$ area centered at Svanhovd. MODIS data indicate a higher photosynthetic activity in 2018 than in 2019. As shown in Fig.~\ref{fig:modis_Psn} we fitted a second order polynomial function of the general form
%
\begin{equation}
A_\mathrm{net}(t) =  -m_0\cdot(t-m_1)^2+m_2,
\end{equation}
%
with the form parameters $m_i$ through the data. Numerically, we retrieved the roots as $G_\mathrm{start}$ 122/106 and $G_\mathrm{end}$ 261/274 for 2018 and 2019, respectively.

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV contributed to framing this research article, has performed the on-site observation of vegetation damage induced by ozone, and advising in plant physiological processes. LE has contributed to framing of the this article, $\mathrm{DO_3SE}$ modeling, and development of the PDF-based temperature acclimation. CO has collected species or PFT parameters from literature, performed all $\mathrm{DO_3SE}$ simulations and initial validation of the results. AE contributed with her experience regarding subarctic vegetation in Finnmark. FS contributed with an assessment of the 2018 meteorological conditions. TB gave guidance in a broader research sense. All authors contributed to discussions, writing, and conclusions.} %% this section is mandatory
